% -*- fill-column: 80 -*-

\documentclass[sigplan, nonacm, natbib = false, review, screen]{acmart}

\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  columns=fullflexible,
  keepspaces=true
}

\lstdefinestyle{metl}{
  language=[Objective]Caml,
  morekeywords={eff, handle, end, return, mask, forall, Effect, alias},
  morecomment=[l]{--},
  literate=
    {->}{{$\to$}}2
    {=>}{{$\Rightarrow$}}2
    {()}{{()}}2
}

\lstdefinestyle{koka}{
  morekeywords={fun,rec,handler,forall},
  sensitive=true,
  morecomment=[l]{--},
  morestring=[b]",
  literate=
    {->}{{$\to$}}2
    {()}{{()}}2
}

\usepackage[backend=bibtex, style=numeric]{biblatex}
\addbibresource{abstract.bib}
\renewcommand*{\bibfont}{\footnotesize}

\usepackage{mathpartir}

%% Ghosts

\usepackage{halloweenmath}
\makeatletter
\newcommand*\forallghostraise{.16ex} % vertical nudge upwards
\newcommand*\forallscale{0.4}       % scale factor for ∀

\newcommand{\forallghost}{\mathord{\mathpalette\forallghost@{}}}
\newcommand{\forallghost@}[2]{%
  \ooalign{%
    % ghost outline
    \hfil$\m@th#1\mathpalette\raiseghost\relax$\hfil\cr
    % smaller forall symbol, raised
    \hfil\raise\forallghostraise\hbox{%
      \scalebox{\forallscale}{$\m@th#1\forall$}}%
    \hfil\cr
  }%
}

\makeatother


\newcommand{\myghost}{\mathpalette\raiseghost\relax}
\newcommand{\raiseghost}[2]{\raisebox{.5\depth}{\scalebox{0.9}{$#1\mathghost$}}}

\newcommand{\Hole}{{\myghost}}
\newcommand{\TopHole}{\forallghost} % implicit forall

%%

%% Keywords

\newcommand{\keyw}[1]{{\textbf{{\textsf{#1}}}}}
\newcommand{\Do}{\keyw{do}\;}

\usepackage{xspace}
\newcommand{\Met}{\textsc{Met}\xspace}
\newcommand{\Metl}{\textsc{Metl}\xspace}
%% AB: probably change this name
\newcommand{\MetlFrost}{\text{Fresco-}\textsc{Metl}\xspace}
\newcommand{\Frost}{\text{Fresco}\xspace}
\newcommand{\MetlF}{\text{F-}\textsc{Metl}\xspace}
%%

%% Margin notes.
%% For submission, comment all the following lines out ...
\paperwidth    =\dimexpr\paperwidth     + 4cm\relax
\oddsidemargin =\dimexpr\oddsidemargin  + 2cm\relax
\evensidemargin=\dimexpr\evensidemargin + 2cm\relax
\marginparwidth=\dimexpr\marginparwidth + 2.2cm\relax
\usepackage{todonotes,setspace}
\newcommand{\todobox}[3]{{\todo[color=#1!20,bordercolor=#1!80]{%
  \begin{spacing}{0.9}\footnotesize\textbf{#2:} #3\end{spacing}}}}
\newcommand{\wenhao}[1]{\todobox{blue}{Wenhao}{#1}}
\newcommand{\aghilas}[1]{\todobox{red}{Aghilas}{#1}}
\newcommand{\sam}[1]{\todobox{yellow}{Sam}{#1}}
%% ... up to here

\title{Practical programming with Modal Effect Types}
%% SL: we can save some space by getting rid of this subtitle
\subtitle{Research Presentation}
\settopmatter{authorsperrow=3}

\author{Aghilas Y. Boussaa}
\orcid{0009-0005-5954-1711}
\email{aghilas.boussaa@ens.fr}
\affiliation{
  \institution{École normale supérieure -- PSL}
  \country{France}
}

\author{Wenhao Tang}
\orcid{0009-0000-6589-3821}
\email{wenhao.tang@ed.ac.uk}
\affiliation{%
  \institution{The University of Edinburgh}
  \country{UK}
}

%% \author{Shengyi Jiang}
%% \orcid{0000-0002-4443-0753}
%% \email{syjiang@cs.hku.hk}
%% \affiliation{%
%%   \institution{The University of Hong Kong}
%%   \country{China}
%% }

%% \author{Bruno C. d. S. Oliveira}
%% \orcid{0000-0002-1846-7210}
%% \email{bruno@cs.hku.hk}
%% \affiliation{%
%%   \institution{The University of Hong Kong}
%%   \country{China}
%% }

\author{Sam Lindley}
\orcid{0000-0002-1360-4714}
\email{sam.lindley@ed.ac.uk}
\affiliation{%
  \institution{The University of Edinburgh}
  \country{UK}
}

\begin{document}
\sloppy
\begin{abstract}
  % Effect handlers are a powerful programming construct allowing users
  % to define and combine computational effects.
  %
  Effect systems statically track the effects that a program can perform.
%
  Traditional effect systems rely on effect variables to provide modular type
  signatures for higher-order functions.
%
  Modal effect types use modalities to give modular and succinct types to
  higher-order functions, reducing the need for effect variables meanwhile still
  allowing higher-order functions to be instantiated with different effects.
%
  However, the practical implementation of modal effect types, and in particular
  how to infer modalities, remains largely unexplored.

  We outline our implementation of a toy ML-like programming language with modal
  effect types.
%
  Our main novelty is that we extend \Frost, a novel bidirectionally-typed
  calculus of first-class polymorphism, to infer first-class modalities.
\end{abstract}
\maketitle
\section{Introduction}
\subsection{Existing type-and-effect systems}
Traditional languages with effect type systems, such as Koka~\cite{koka2014},
annotate each arrow with the effects a function of this type may
perform. Consequently, each function inherits, in its type signature, the
effects of its callees. This becomes cumbersome when typing higher-order
functions, and is an obstacle to backward compatibility (adding effect typing to
an existing language requires changing the type signatures of all higher-order
functions). For instance (annotating arrows with effects in square brackets),
the function \lstinline{map} has the following type, even though it performs no
effects directly.
\begin{lstlisting}[style=metl]
  forall e . (list a, a ->[e] b) ->[e] list b
\end{lstlisting}
The effect variable \lstinline{e} allows \lstinline{map} to take any effectful
function as an argument.
%
Syntactic sugar can simplify such signatures, as demonstrated by
Frank~\cite{frank2020}. But relying on syntax makes the system fragile.
%
The language Effekt~\cite{effekt2020} mitigates this problem with a
\emph{capability} based system. However, to guarantee effect-safety, functions
become second-class objects, which hinders backwards compatibility.

%% SL: I think when we mention ``backwards compatibility'' here we're alluding
%% to the idea that one might wish to extend an existing language with support
%% for effect typing. But we haven't spelled this out here, so it will likely be
%% confusing to readers.
%
%% SL: I've added a parenthetical explanation of this


\subsection{Modal Effect Types}

%% \sam{Introduce the \Met calculus!}

%% \sam{We need to explain \emph{how} the modality $[]$ modifies the
%%   ambient effect context here - and in particular how conceptually it
%%   applies to all of the arrows contained in its scope.}

Modal effect types remove the need for effect variables by not tying
effects to arrows ~\cite{tang2025modal, tang2026modal}.
%
An \emph{effect context} denotes the effects that a computation can perform.
%
A \emph{modality} is a transformation on effect contexts.
%% , typically the
%% currently active effect context which we refer to as the \emph{ambient effect
%% context}.
%
%% %draws inspiration from Multimodal Type Theory~\cite{mtt2020} and
%% \emph{effect contexts} represent the effects a computation can perform;
%% \emph{modalities} then modify effect contexts.
%
The \Met core calculus implements these ideas. It is an extension of System F
that adds effect handlers and modalities.
%
Modalities may appear directly in types such as $\mu A$, where $\mu$ is a
modality and $A$ is a type. A value of type $\mu A$ is a value of type $A$ whose
effects are those of the ambient effect context transformed by the modality
$\mu$. A modality acts hereditarily on every arrow in $A$.
%
For instance, the \textsf{map} function has the following type, where $[]$ is a
modality that indicates that the function does not perform any effect.
\[
  \forall a b . []((a \to b) \to \textsf{list } a \to \textsf{list } b)
\]
In general, $[E]$ denotes an \emph{absolute} modality, which overwrites the
ambient effects with $E$.
%
Replacing the ambient effect context by the empty one may seem overly
restrictive. However, when we instantiate \textsf{map} and eliminate the
modality we may soundly upcast the empty effect context to whatever the ambient
effect context happens to be.
%
%% In general, $[E]$ denotes an \emph{absolute} modality, which overwrites the
%% input effect context with $E$.
%% %
%% If we write $[E]A$ then $A$ asserts
%
%% In this case $E$ is empty.
%% %
%
%% , which means that it replaces
%% the input effect context with a
%
%% the ambient
%% effect context with an independent one; here the empty context. In general, the
%% modality $[E]$ replaces the ambient effect context with $E$. In the example
%% above, the modality implies that if any of the arrow in the type of \textsf{map}
%% performs an effect, then this effect comes from the ambient context of one of
%% the parent computations.
%
\Met also provides another kind of modality: a \emph{relative} modality $\langle
L|D\rangle$ removes the effects in $L$ from the ambient effect context before
adding the effects in $D$. Relative modalities are important for typing effect
handlers.

\Met provides explicit constructs for introducing and eliminating modalities.
%
The $\keyw{mod}_\mu$ and $\keyw{let mod}_\mu$ constructs respectively introduce
and eliminate modality $\mu$. For example, the following term upcasts a pure
function $x:[](\textsf{int} \to\textsf{int})$:
\[
\left(\keyw{let mod}_{[]} y = x \keyw{ in } \keyw{mod}_{[E]} y\right) : [E](\textsf{int} \to\textsf{int})
\]
%
%% SL: I'm not sure that the following really adds much, so I've commented it
%% out for now.
%
%% The $\keyw{mod}_\mu V$ expression modifies the effect context of the value $V$
%% by applying modality $\mu$. Let-bindings allow us to retrieve the original
%% value: $\keyw{let mod}_\mu x = \keyw{mod}_\mu V \keyw{ in } x$ reduces to $V$.
%
See~\cite{tang2025modal, tang2026modal} for a complete introduction to modal
effect types.


Just as writing explicit type abstractions and type applications in programs is
inconvenient, writing explicit introduction and elimination forms for modalities
is too.
%
The original \Met paper~\cite{tang2025modal} provides a bidirectional algorithm
that translates a surface language called \Metl into \Met; to dinstiguish it
from the languages we will introduce later, we will refer to the original \Metl
as \MetlF. But this algorithm only infers modalities: \MetlF still requires
explicit type applications for value polymorphism.
%
A central aim of our work is to combine modality inference with a bidirectional
typing algorithm for first-class polymorphism.

% AB: removed the paragraph
%\sam{I'm not clear on when, where or why we need a type annotation
%  here.}

\subsection{\Frost type inference}
The usual way to type check an application with a bidirectional type system is
to first infer the type of the function and then use this type to check the
argument~\cite{dunfield2013complete, greedy2023, jiang2025bidirectional}. Less
commonly, some type systems allow information to flow from the argument to the
function~\cite{pierce2000local, xie2018let}.

\Frost supports information flow in both directions~\cite{tang2025ML}. To do so,
the type system relies on \emph{skeletons}. Skeletons are types augmented with
\emph{ghosts} ($\Hole$) and \emph{universal ghosts} ($\TopHole$). Ghosts and
universal ghosts amount to special forms of holes which appear in type or
quantifier position respectively. The typing rules for applications infers a
skeleton for the argument, uses the skeleton to infer a type for the function,
then checks the argument against the function type's argument. The skeleton
allows information flow from argument to function before the flow is reversed.

For instance, consider $\textsf{poly'}:\forall a.(\forall b. b\to b\times a)\to
a$ and $\textsf{ids}:\textsf{list }(\forall a.a\to a)$. \Frost infers the type
$\textsf{list }(\forall a.a\to a)$ for the expression $\textsf{poly'} (\lambda
x. (x, \textsf{ids}))$.
%
First, the skeleton $\TopHole.\Hole\to\Hole\times(\TopHole.\textsf{list
}(\forall a.a\to a))$ of the argument instantiates the function's type variable
$a$ to $\textsf{list }(\forall a.a\to a)$.
%
Then, checking the argument against $\forall b.b\to b\times\textsf{list
}(\forall a.a\to a)$, wraps the argument in a type abstraction.

As noted in previous work on \Met, elimination and introduction of modalities is
reminiscent of instantiation and generalisation in type inference~\cite[Section
  7]{tang2025modal}. Pursuing this analogy, universal ghosts allow \Frost to
treat quantifiers and modalities with similar rules.

We will now outline \MetlFrost, a toy programming language which extends \Frost
to infer modal effect types. We use examples in \Met, \MetlF and \MetlFrost to
evaluate the capabilities of the latter and discuss limitations. More examples,
as well as a test suite are available
\href{https://github.com/4y8/modal-effect-types/tree/frost}{online}.

\section{Programming with Modal Effect Types}
\subsection{Our implementation}

\sam{We haven't said what effect structures are. I suspect it may be best just
  to avoid mentioning them in this extended abstract.}

The core language of our implementation is \Met with scoped
rows~\cite{scopedrows2005} as an effect structure (i.e.\ modalities will have
scoped rows of effects) with additions such as parametric effects and algebraic
data types.

We have implemented three surface languages for this extension of \Met, which
all support modality inference: a version with explicit type application, one
with type inference for types without modalities or quantifiers based on an
algorithm of Dunfield and Krishnaswami~\cite{dunfield2013complete}, and one with
type inference based on \Frost. They are all available
\href{https://github.com/4y8/modal-effect-types/}{online} and the latter also
has a \href{https://www.normalesup.org/~boussaa/met/playground/}{web
  interface}. The web page features several effectful programs as examples. They
all require only top-level type annotations to type check, offering some
evidence that \MetlFrost is a usable surface language for programming with modal
effect types.

\sam{In this section we seem to have started talking about boxing and
  unboxing. Previously we only used introduction and elimination of modalities.}

The new subtyping and unification rules for modalities are much like those for
universal quantification. However, they are more involved as modalities hold
more information. For instance, unification of two modal types also unifies the
two top-level modalities.

%% \aghilas{I am not sure of how to integrate these paragraphs, if they are
%%   relevant}

\sam{The following paragraph is perhaps still not entirely accurate.}
%
\Met has modality-annotated variable declarations in its contexts:
$x:_\mu A$, where $A$ can also have a modality. For uniformity, \MetlFrost, like
\MetlF, has plain bindings $x:\mu G$, where $G$ is a type without a top-level
modality, and implicitly unboxes all modalities on variable types ($x:\mu G$
represents $x:_\mu G$).
%
%% If $G$ has no top-level modality, then we interpret this binding
%% as $x:_\mu G$ in \Met.

%% arrays
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}

%% lists
\newcommand{\bl}{\ba[t]{@{}l@{}}}
\newcommand{\el}{\ea}

\subsection{Iterating on a list}
The standard iter higher-order function has type $\forall a . []((a \to
\textsf{unit})\to \textsf{unit})$ in \Met. To call itself, the function unboxes
the absolute modality of $\textsf{iter}\, @a$, where the at sign indicates type
application.
%
\aghilas{not sure about the syntax we should use here}
%
\sam{I think this syntax is probably fine, but we should probably say something
  about switching to an ASCII syntax for the surface languages.}
%
\[
\bl
\Lambda a^{\textsf{Any}}.\,
  \bl
  \keyw{mod}_{[]}.\lambda f^{a\to\textsf{unit}}\,l^{\textsf{list }a}. \\
  \keyw{match}\; l \;\keyw{with} \\
  \quad
    \bl
    \mid \textsf{Nil} \to () \\
    \mid \textsf{Cons}(hd,tl) \to
    \keyw{let\;mod}^{\langle\rangle}\_ = f\,hd \;\keyw{in} \\
    \qquad
       \bl
       \keyw{let\;mod}^{[]}\; i = \textsf{iter}\, @a \;\keyw{in}\; \\
        i\,f\,tl
       \el
    \el
  \el
\el
\]
%% \[
%% \Lambda a^{\textsf{Any}}.\keyw{mod}_{[]}\lambda f^{a\to\textsf{unit}} l^{\textsf{list }a}.\keyw{match } l \keyw{ with }\textsf{Nil}\to () \mid \textsf{Cons}(hd, tl)\to
%% \keyw{let mod}^{\langle\rangle}\_ = f\; hd \keyw{ in }
%% \keyw{let mod}^{[]}i = \textsf{iter}\; \{ a \} \keyw{ in }
%% i\; f\; tl
%% \]

The type for \MetlF and \MetlFrost is the same as the \Met one.
%
\MetlF avoids explicit modality elimination.
\begin{lstlisting}[style=metl]
  val iter : forall a . []((a -> unit) -> unit)
  let iter f l =
    match l with
    | Nil -> ()
    | Cons (hd, tl) -> f hd; iter @a f tl
    end
\end{lstlisting}
\MetlFrost removes the type application.
\begin{lstlisting}[style=metl]
    | Cons (hd, tl) -> f hd; iter f tl
\end{lstlisting}

\subsection{Searching a list with effects}
Calling the iterator from the last subsection with an effectful function as
argument can search a list, thanks to the parametric effect below.
\begin{lstlisting}[style=metl]
  eff gen [a] = yield : a => unit
\end{lstlisting}
The declaration above defines a parametric effect \lstinline[style=metl]{gen}
with a single operation \lstinline[style=metl]{yield}. The brackets around the
parameter \lstinline[style=metl]{a} indicate that it has kind \textsf{Abs}: the
kind of types whose meaning is independent from the ambient effect
context. There is another kind, called \textsf{Any}, that includes all types. To
avoid effect leakage, the types of inputs and outputs of operations have kind
\textsf{Abs}.

Once again, \MetlF requires no explicit modality introduction or elimination.
\begin{lstlisting}[style=metl]
  val find :
    forall [a] . []((a -> bool) -> list a -> option a)
  let find p l =
    handle<gen a> iter @a
      (fun x -> if mask<gen> (p x) then do yield x else ())
      l with
    | return _ => None
    | yield x k => Some (x)
    end
\end{lstlisting}
The angle brackets that follow the \lstinline[style=metl]{handle} keyword
provide a parameter for the effect \lstinline[style=metl]{gen}. \Frost removes
the need for this annotation, as well as for the explicit type application.

\subsection{Some limitations}
\paragraph{Syntactic treatment of quantifiers}
Our current unification rules have a syntactic treatment of quantifiers. We can
unify neither $[l]\langle l'\rangle(\textsf{unit}\to\textsf{unit})$ and $[l',
  l](\textsf{unit}\to\textsf{unit})$, nor $\forall a b . a \times b$ and
$\forall b a . a \times b$, though they are each subtypes of one another.

\paragraph{Lack of kind inference}
The type inference rule for lambda abstractions introduces a fresh type variable
for the argument. As the system lacks kind inference, the kind of this variable
is \textsf{Any} to be permissive. Consequently, functions that take pure values
require annotation. For instance, the \lstinline[style=metl]{find} function
above does not type check without the top-level annotation because the variable
\lstinline[style=metl]{x} is not pure.

\sam{This explanation is too fuzzy. I don't know what aspect of the type
  annotation it refers to or what the variable x has to do with it.}

\subsection{Conclusion and future work}
The extension of \Frost to modal effect types is a promising approach to
integrating effect handlers into functional programming languages.
%
Future work includes taking effect contexts into account in skeleton inference,
extending \Frost to support a fragment of System $F^\omega$, proving the
soundness of the \Frost extension, and exploring a more semantic treatment of
quantifiers and modalities.

\printbibliography

\end{document}

% LocalWords:  polymorphism Koka callees forall effectful Effekt iter
% LocalWords:  subeffecting instantiation hd tl bool
% LocalWords:  metl subtype
